\section*{Supplementary Exercises for Chapter~\ref{chap:4}}
\addcontentsline{toc}{section}{Supplementary Exercises for Chapter~\ref{chap:4}}

\begin{Filesave}{solutions}
\solsection{Supplementary Exercises for Chapter~\ref{chap:4}}
\end{Filesave}

\begin{multicols}{2}
\begin{supex} \label{supex:sup_ch4_ex1}
Suppose that $\vect{u}$ and $\vect{v}$ are nonzero vectors. If $\vect{u}$ and $\vect{v}$ are not parallel, and $a\vect{u} + b\vect{v} = a_{1}\vect{u} + b_{1}\vect{v}$, show that $a = a_{1}$ and $b = b_{1}$.

\end{supex}

\begin{supex}
Consider a triangle with vertices $A$, $B$, and $C$. Let $E$ and $F$ be the midpoints of sides $AB$ and $AC$, respectively, and let the medians $EC$ and $FB$ meet at $O$. Write $\longvect{EO} = s\longvect{EC}$ and $\longvect{FO} = t\longvect{FB}$, where $s$ and $t$ are scalars. Show that $s = t = \frac{1}{3}$ by expressing $\longvect{AO}$ two ways in the form $a\longvect{EO} + b\longvect{AC}$, and applying Exercise \ref{supex:sup_ch4_ex1}. Conclude that the medians of a triangle meet at the point on each that is one-third of the way from the midpoint to the vertex (and so are concurrent).

\end{supex}

\begin{supex}
A river flows at $1$ km/h and a swimmer moves at $2$ km/h (relative to the water). At what angle must he swim to go straight across? What is his resulting speed?

\end{supex}

\begin{supex}
A wind is blowing from the south at $75$ knots, and an airplane flies heading east at $100$ knots. Find the resulting velocity of the airplane.

\begin{supsol}
$125$ knots in a direction $\theta$ degrees east of north, where $\cos\theta = 0.6$ ($\theta = 53^\circ$  or $0.93$ radians).

\end{supsol}
\end{supex}

\begin{supex}
An airplane pilot flies at $300$ km/h in a direction $30^\circ$ south of east. The wind is blowing from the south at $150$ km/h.


\begin{enumerate}[label={\alph*.}]
\item Find the resulting direction and speed of the airplane.

\item Find the speed of the airplane if the wind is from the west (at $150$ km/h).

\end{enumerate}
\end{supex}

\begin{supex}
A rescue boat has a top speed of $13$ knots. The captain wants to go due east as fast as possible in water with a current of $5$ knots due south. 
Find the velocity vector $\vect{v} = (x, y)$ that she must achieve, assuming the $x$ and $y$ axes point east and north, respectively, and find her resulting speed.

\begin{supsol}
$(12, 5)$. Actual speed $12$ knots.

\end{supsol}
\end{supex}


\begin{supex}
A boat goes $12$ knots heading north. The current is $5$ knots from the west. In what direction does the boat actually move and at what speed?

\end{supex}

\begin{supex}
Show that the distance from a point $A$ (with vector $\vect{a}$) to the plane with vector equation $\vect{n} \dotprod \vect{p} = d$ is $\frac{1}{\vectlength n \vectlength}|\vect{n} \dotprod \vect{a} - d|$.

\end{supex}

\begin{supex}
If two distinct points lie in a plane, show that the line through these points is contained in the plane.

\end{supex}



\begin{supex}
The line through a vertex of a triangle, perpendicular to the opposite side, is called an \textbf{altitude}\index{altitude}\index{triangle!altitude} of the triangle. Show that the three altitudes of any triangle are 
concurrent. (The intersection of the altitudes is called the \textbf{orthocentre}\index{triangle!orthocentre}\index{orthocentre} of the triangle.) [\textit{Hint}: If $P$ is the intersection of two of the altitudes, show that the line through $P$ and the remaining vertex is perpendicular to the remaining side.]

\end{supex}
\end{multicols}
