\section*{Exercises for \ref{sec:2_8}}

\begin{Filesave}{solutions}
\solsection{Section~\ref{sec:2_8}}
\end{Filesave}

\begin{multicols}{2}
\vspace*{-2em}
\begin{ex}
Find the possible equilibrium price structures when the input-output matrices are:
\begin{exenumerate}
\exitem $\leftB \begin{array}{rrr}
0.1 & 0.2 & 0.3 \\
0.6 & 0.2 & 0.3 \\
0.3 & 0.6 & 0.4
\end{array} \rightB$
\exitem $\leftB \begin{array}{rrr}
0.5 & 0 & 0.5 \\
0.1 & 0.9 & 0.2 \\
0.4 & 0.1 & 0.3
\end{array} \rightB$
\exitem* $\leftB \begin{array}{rrrr}
0.3 & 0.1 & 0.1 & 0.2 \\
0.2 & 0.3 & 0.1 & 0 \\
0.3 & 0.3 & 0.2 & 0.3 \\
0.2 & 0.3 & 0.6 & 0.5
\end{array} \rightB$
\exitem* $\leftB \begin{array}{rrrr}
0.5 & 0 & 0.1 & 0.1 \\
0.2 & 0.7 & 0 & 0.1 \\
0.1 & 0.2 & 0.8 & 0.2 \\
0.2 & 0.1 & 0.1 & 0.6
\end{array} \rightB$
\end{exenumerate}
\begin{sol}
\begin{enumerate}[label={\alph*.}]
\setcounter{enumi}{1}
\item $\leftB \begin{array}{r}
t \\
3t \\
t
\end{array} \rightB$

\setcounter{enumi}{3}
\item $\leftB \begin{array}{r}
14t \\
17t \\
47t \\
23t
\end{array} \rightB$

\end{enumerate}
\end{sol}
\end{ex}

\begin{ex}
Three industries $A$, $B$, and $C$ are such that all the output of $A$ is used by $B$, all the output of $B$ is used by $C$, and all the output of $C$ is used by $A$. Find the possible equilibrium price structures.

\begin{sol}
$\leftB \begin{array}{r}
t \\
t \\
t
\end{array} \rightB$
\end{sol}
\end{ex}

\begin{ex}
Find the possible equilibrium price structures for three industries where the input-output matrix is $\leftB \begin{array}{rrr}
1 & 0 & 0 \\
0 & 0 & 1 \\
0 & 1 & 0
\end{array} \rightB$. Discuss why there are two parameters here.
\end{ex}

\begin{ex}
Prove Theorem~\ref{thm:007013} for a $2 \times 2$ stochastic matrix $E$ by first writing it in the form $E = \leftB \begin{array}{cc}
a & b \\
1 - a & 1 - b
\end{array} \rightB$, where $0 \leq a \leq 1$ and $0 \leq b \leq 1$.

\begin{sol}
$P = \leftB \begin{array}{c}
bt \\
(1 - a)t
\end{array} \rightB$
 is nonzero (for some $t$) unless $b = 0$ and $a = 1$. In that case, $\leftB \begin{array}{r}
 1 \\
 1
 \end{array} \rightB$
 is a solution. If the entries of $E$ are positive, then $P = \leftB \begin{array}{c}
 b \\
 1 - a
 \end{array} \rightB$
 has positive entries.
\end{sol}
\end{ex}

\begin{ex}
If $E$ is an $n \times n$ stochastic matrix and $\vect{c}$ is an $n \times 1$ matrix, show that the sum of the entries of $\vect{c}$ equals the sum of the entries of the $n \times 1$ matrix $E\vect{c}$.
\end{ex}

\begin{ex}
Let $W = \leftB \begin{array}{ccccc}
1 & 1 & 1 & \cdots & 1
\end{array} \rightB$. Let $E$ and $F$ denote $n \times n$ matrices with nonnegative entries.

\begin{enumerate}[label={\alph*.}]
\item Show that $E$ is a stochastic matrix if and only if $WE = W$.

\item Use part (a.) to deduce that, if $E$ and $F$ are both stochastic matrices, then $EF$ is also stochastic.

\end{enumerate}
\end{ex}

\begin{ex}
Find a $2 \times 2$ matrix $E$ with entries between $0$ and $1$ such that:

\begin{enumerate}[label={\alph*.}]
\item $I - E$ has no inverse.

\item $I - E$ has an inverse but not all entries of $(I - E)^{-1}$ are nonnegative.

\end{enumerate}
\begin{sol}
\begin{enumerate}[label={\alph*.}]
\setcounter{enumi}{1}
\item $\leftB \begin{array}{rr}
0.4 & 0.8 \\
0.7 & 0.2
\end{array} \rightB$

\end{enumerate}
\end{sol}
\end{ex}

\begin{ex}
If $E$ is a $2 \times 2$ matrix with entries between $0$ and $1$, show that $I - E$ is invertible and $(I - E)^{-1} \geq 0$ if and only if $\func{tr }E < 1 + \func{det }E$. Here, if $E = \leftB \begin{array}{rr}
a & b \\
c & d
\end{array} \rightB$,
 then $\func{tr }E = a + d$ and $\func{det }E = ad - bc$.

\begin{sol}
If $E = \leftB \begin{array}{rr}
	a & b \\
	c & d
\end{array} \rightB$,
 then $I - E = \leftB \begin{array}{cc}
 1 - a & -b \\
 -c & 1 - d
 \end{array} \rightB$,
 so $\func{det}(I - E) = (1 - a)(1 - d) - bc = 1 - \func{tr }E + \func{det }E$. If $\func{det}(I - E) \neq 0$, then $(I - E)^{-1} = \frac{1}{\func{det}(I - E)} \leftB \begin{array}{cc}
 1 - d & b \\
 c & 1 - a
 \end{array} \rightB$, so $(I - E)^{-1} \geq 0$ if $\func{det}(I - E) > 0$, that is, $\func{tr }E < 1 + \func{det } E$.
 The converse is now clear.
\end{sol}
\end{ex}

\begin{ex}
In each case show that $I - E$ is invertible and $(I - E)^{-1} \geq 0$.
\begin{exenumerate}
\exitem $\leftB \begin{array}{rrr}
0.6 & 0.5 & 0.1 \\
0.1 & 0.3 & 0.3 \\
0.2 & 0.1 & 0.4
\end{array} \rightB$
\exitem $\leftB \begin{array}{rrr}
0.7 & 0.1 & 0.3 \\
0.2 & 0.5 & 0.2 \\
0.1 & 0.1 & 0.4
\end{array} \rightB$
\exitem $\leftB \begin{array}{rrr}
0.6 & 0.2 & 0.1 \\
0.3 & 0.4 & 0.2 \\
0.2 & 0.5 & 0.1
\end{array} \rightB$
\exitem $\leftB \begin{array}{rrr}
0.8 & 0.1 & 0.1 \\
0.3 & 0.1 & 0.2 \\
0.3 & 0.3 & 0.2
\end{array} \rightB$
\end{exenumerate}
\begin{sol}
\begin{enumerate}[label={\alph*.}]
\setcounter{enumi}{1}
\item  Use $\vect{p} = \leftB \begin{array}{r}
3 \\
2 \\
1
\end{array} \rightB$
 in Theorem~\ref{thm:007060}.

\setcounter{enumi}{3}
\item  $\vect{p} = \leftB \begin{array}{r}
3 \\
2 \\
2
\end{array} \rightB$
 in Theorem~\ref{thm:007060}.

\end{enumerate}
\end{sol}
\end{ex}

\begin{ex}\label{ex:ex2_8_10}
Prove that (1) implies (2) in the Corollary to Theorem~\ref{thm:007060}.
\end{ex}

\begin{ex}\label{ex:ex2_8_11}
If $(I - E)^{-1} \geq 0$, find $\vect{p} > 0$ such that $\vect{p} > E\vect{p}$.
\end{ex}

\begin{ex}\label{ex:ex2_8_12}
If $E\vect{p} < \vect{p}$ where $E \geq 0$ and $\vect{p} > 0$, find a number $\mu$ such that $E\vect{p} < \mu\vect{p}$ and $0 < \mu < 1$.


[\textit{Hint}: If $E\vect{p} = (q_{1}, \dots, q_{n})^{T}$ and $\vect{p} = (p_{1}, \dots, p_{n})^{T}$, take any number $\mu$ where $\func{max}\left\lbrace \frac{q_{1}}{p_{1}}, \dots, \frac{q_{n}}{p_{n}} \right\rbrace < \mu < 1$.]
\end{ex}

\end{multicols}
